---
title: MSc Health Data Science Project Title
author:
  - name: Manasvi Saite
  - name: Kaushal Mayur
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: "/Users/manasvi15/Downloads/Project/UOG_logo.png"

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
primary_colour:	"#a80050"	# Secondary colour use for poster design.
secondary_colour:	"#84003d"	# Main colour used for poster design.
accent_colour:	"#e6007e"	# A third colour option.
author_textcol: "white"
title_textsize: "80pt"
author_textsize: "50pt"
body_textsize: "20pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
a, .section, p {
  overflow-wrap: anywhere;
  word-break: break-word;
}
.section h4 { break-after: column; }
div.footnotes { font-size: 18pt; }
/* Add space between table and next plot */
div.chunk { margin-bottom: 20px; }
table { margin-bottom: 25px !important; }

```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

options(knitr.table.format = "html")

library(tidyverse)
library(kableExtra)
library(scales)
library(VGAM)
library(NHANES)
library(plotly)
```

# Background and Problem

Healthcare analytics often requires models suited to non‑standard outcomes (e.g., counts or ordered categories), where standard linear regression is inappropriate.  
In this poster we study **self‑reported general health** (HealthGen) from NHANES, recorded on an ordered scale from "Excellent" to "Poor", and use an ordinal regression framework to quantify how age, BMI, sex and smoking shift health ratings.

## Objectives of Project
- Fit a proportional‑odds (cumulative logit) model for HealthGen.  
- Estimate and visualise predicted probabilities of each health category across age.  
- Compare predicted health profiles for current smokers vs non‑smokers.

# Data Sources and Datasets


**NHANES Overview**: Nationally representative survey of ~10,000 US residents annually. Combines interviews, physical exams, and lab tests from civilian noninstitutionalized population. [web:297]

**Dataset**: NHANES 2013-2014 (R package `NHANES`). ~9,700 participants. [web:303]

**HealthGen (Outcome)**: Self-reported general health rating (ages ≥12). 5 ordered categories:
- **Excellent** (best) → **Very good** → **Good** → **Fair** → **Poor** (worst)
- Typical US distribution: 15% Excellent, 41% Very good, 49% Good, 16% Fair, 4% Poor [web:317]

**Key Predictors** (supervisor-specified model):
- **Age**: Years (continuous, adults only)
- **Gender**: Male/Female
- **BMI**: kg/m² (body mass index from measured height/weight)
- **SmokeNow**: Current smoker (Yes/No)

**Analytic Sample**: Complete cases only (n=5,736). Filtered for non-missing Age, Gender, BMI, SmokeNow, HealthGen.

<!-- This #### below causes the columns to break -->
```{r data_load, include=FALSE}
library(NHANES)
data(NHANES)
nhanes_raw <- NHANES::NHANES

nhanes_model <- nhanes_raw %>%
  mutate(
    GENHLTH = ordered(HealthGen, levels = c("Excellent","Very good","Good","Fair","Poor")),
    Gender  = factor(Gender),
    SmokeNow = factor(SmokeNow, levels = c("No","Yes"))
  ) %>%
  filter(complete.cases(GENHLTH, Age, Gender, BMI, SmokeNow)) %>%
  mutate(GENHLTH = droplevels(GENHLTH))

fit <- VGAM::vglm(
  GENHLTH ~ Age + Gender + BMI + SmokeNow,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE),
  data = nhanes_model
)

Mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

health_colors <- c(
  "Excellent"="red3","Very good"="green4","Good"="cyan3","Fair"="green3","Poor"="purple3"
)

```
<!-- You can change where you put these breaks, and the "column_numbers: 4" argument in the YAMLS header to change the layout of the poster, with care! -->

#### 

# Early Results / Descriptive Statistics of Datasets

The modelling dataset includes **`r format(nrow(nhanes_model), big.mark=",")`** complete cases with mean age
**`r sprintf("%d±%d", round(mean(nhanes_model$Age)), round(sd(nhanes_model$Age)))`** years and mean BMI
**`r sprintf("%.1f±%.1f", mean(nhanes_model$BMI), sd(nhanes_model$BMI))`** kg/m².

Current smoking prevalence is **`r scales::percent(mean(nhanes_model$SmokeNow=="Yes"), accuracy=0.1)`**
and females comprise **`r scales::percent(mean(nhanes_model$Gender=="female"), accuracy=0.1)`** of the analytic sample.


Some basic summaries of the dataset are below:

```{r mytable, out.width='85%', fig.show='hold', tab.cap="Table of Nhanes data", tab.height = 6}
tab1 <- tibble(
  Measure = c("N participants", "Age (mean ± SD)", "BMI (mean ± SD)", "Current smokers (%)", "Females (%)"),
  Value = c(
    format(nrow(nhanes_model), big.mark = ","),
    sprintf("%d ± %d", round(mean(nhanes_model$Age)), round(sd(nhanes_model$Age))),
    sprintf("%.1f ± %.1f", mean(nhanes_model$BMI), sd(nhanes_model$BMI)),
    scales::percent(mean(nhanes_model$SmokeNow == "Yes"), accuracy = 0.1),
    scales::percent(mean(nhanes_model$Gender == "female"), accuracy = 0.1)
  )
)

knitr::kable(tab1, col.names = c("Measure", "Value")) %>%
  kable_styling(font_size = 22, full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#a80050", color = "white")

```



```{r figure-1, out.width='92%', fig.height=6, fig.cap='Predicted health probabilities by age (typical covariates).'}
levs <- levels(nhanes_model$GENHLTH)
col_pal <- health_colors[levs]

newdat <- nhanes_model %>%
  summarise(
    Gender = Mode(Gender),
    BMI = mean(BMI, na.rm = TRUE),
    SmokeNow = Mode(SmokeNow)
  ) %>%
  tidyr::crossing(Age = seq(18, 80, length.out = 100))

probs <- as.data.frame(predict(fit, newdat, type = "response"), check.names = FALSE)

# Align to however many columns VGAM returns
levs_use <- levs[seq_len(min(length(levs), ncol(probs)))]
names(probs) <- levs_use

pred <- dplyr::bind_cols(newdat, probs) %>%
  tidyr::pivot_longer(cols = all_of(levs_use), names_to = "Health", values_to = "Probability") %>%
  mutate(Health = factor(Health, levels = levs_use))

ggplot(pred, aes(x = Age, y = Probability, colour = Health)) +
  geom_line(size = 2.2) +
  scale_colour_manual(values = col_pal[levs_use]) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(20, 80, 20)) +
  labs(x = "Age", y = "Probability", colour = "Health Category",
       title = "Predicted Probability") +
  theme_minimal(base_size = 22) +
  theme(legend.position = "top", legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))


```



```{r , out.width='92%', fig.height=6, fig.cap=' Smoking effect on predicted health probabilities.'}
levs <- levels(nhanes_model$GENHLTH)
col_pal <- health_colors[levs]

base_vals <- nhanes_model %>%
  summarise(Gender = Mode(Gender), BMI = mean(BMI, na.rm = TRUE))

new_smoke <- base_vals %>%
  tidyr::crossing(
    SmokeNow = c("No", "Yes"),
    Age = seq(18, 80, length.out = 100)
  )

probs_smoke <- as.data.frame(predict(fit, new_smoke, type = "response"), check.names = FALSE)

levs_use <- levs[seq_len(min(length(levs), ncol(probs_smoke)))]
names(probs_smoke) <- levs_use

pred_smoke <- dplyr::bind_cols(new_smoke, probs_smoke) %>%
  tidyr::pivot_longer(cols = all_of(levs_use), names_to = "Health", values_to = "Probability") %>%
  mutate(
    Health = factor(Health, levels = levs_use),
    SmokeNow = factor(SmokeNow, levels = c("No","Yes"))
  )

ggplot(pred_smoke, aes(x = Age, y = Probability, colour = Health)) +
  geom_line(size = 2.2) +
  facet_wrap(~ SmokeNow, ncol = 2) +
  scale_colour_manual(values = col_pal[levs_use]) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(20, 80, 20)) +
  labs(x = "Age", y = "Probability", colour = "Health Category",
       title = "Effect of Smoking",
       subtitle = "Cumulative logit (proportional odds)") +
  theme_minimal(base_size = 22) +
  theme(legend.position = "top", legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text = element_text(size = 18, face = "bold"))

```



**Interactive: Smoking effect on predicted general health**
```{r interactiveplot, out.width='92%', fig.height=5.2, fig.cap='Interactive plot (hover to see probabilities) showing predicted HealthGen probabilities by age and current smoking status.'}

p_int <- ggplot(
  pred_smoke,
  aes(
    x = Age, y = Probability, colour = Health, group = Health,
    text = paste0(
      "Age: ", round(Age),
      "<br>SmokeNow: ", SmokeNow,
      "<br>Health: ", Health,
      "<br>Prob: ", scales::percent(Probability, accuracy = 0.1)
    )
  )
) +
  geom_line(linewidth = 2) +
  facet_wrap(~ SmokeNow, ncol = 2) +
  scale_colour_manual(values = health_colors[levels(pred_smoke$Health)]) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.65)) +
  scale_x_continuous(breaks = seq(20, 80, 20)) +
  labs(x = "Age", y = "Probability", colour = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top")

plotly::ggplotly(p_int, tooltip = "text") %>%
  config(displayModeBar = "hover", displaylogo = FALSE)
```

#### 

# Next Project Steps

We plan to conduct further analysis:

-Compare model variants using information criteria (AIC/BIC) and diagnostic plots. 

-Consider additional predictors available in NHANES (e.g., activity, alcohol, comorbidity measures). 

-Report uncertainty (confidence intervals) for key effects and validate predictions on a holdout split.




# GitHub

The code and datasets for this project can be viewed at our GitHub repository here: <https://https://github.com/manasvisaite1504/Count_data_health_Industry-project>



# References

NHANES data package (R).

VGAM cumulative logit modelling.

