---
title: "NHANES Ordinal Regression"
author: "Manasvi Saite & Kaushal Mayur"
affiliation: "University of Galway"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(VGAM)
library(knitr)
library(NHANES)
library(shiny)
library(plotly)
health_order <- c("Excellent", "Very good", "Good", "Fair", "Poor")

```

# Data Preparation

```{r}
data(NHANES)

df <- NHANES %>%
  mutate(
    HealthNum = case_when(
      HealthGen == "Excellent" ~ 1,
      HealthGen == "Vgood" ~ 2,
      HealthGen == "Good" ~ 3,
      HealthGen == "Fair" ~ 4,
      HealthGen == "Poor" ~ 5,
      TRUE ~ NA_real_
    ),
    Health = ordered(HealthNum, levels = c(1, 2, 3, 4, 5)),
    Age = as.numeric(Age),
    Gender = factor(Gender),
    BMI = as.numeric(BMI),
    Smoking = factor(SmokeNow)
  ) %>%
  dplyr::select(Health, Age, Gender, BMI, Smoking) %>%
  na.omit() %>%
  droplevels()

summary(df)
```

# Exploratory Data Analysis

## Colour Names Used in Graphs

```{r colours}
bar_colour <- "pink"
histogram_colour <- "grey"
box_fill_colour <- "pink"
box_border_colour <- "grey"
```

## Missing Data 

```{r}
missing_summary <- data.frame(
  Variable = names(df),
  Missing_Count = sapply(df, function(x) sum(is.na(x))),
  Missing_Percent = round(sapply(df, function(x) mean(is.na(x)) * 100), 2)
)

kable(missing_summary)
```

## Health Count

```{r}
kable(as.data.frame(table(df$Health)), col.names = c("Health", "Count"))
```


## Gender Count 

```{r}
kable(as.data.frame(table(df$Gender)), col.names = c("Gender", "Count"))
```

## Smoking Status Count

```{r}
kable(as.data.frame(table(df$Smoking)), col.names = c("Smoking", "Count"))
```

## Age Summary

```{r}
summary(df$Age)
```

## Bmi Summary

```{r}
summary(df$BMI)
```


## Distribution of Self-Reported Health

```{r}
ggplot(df, aes(x = Health)) +
  geom_bar(fill = bar_colour, colour = "white") +
  labs(
    title = "Distribution of Self-Reported Health",
    x = "Health Category",
    y = "Count"
  ) +
  theme_minimal()
```

## Distribution of Gender

```{r}
ggplot(df, aes(x = Gender)) +
  geom_bar(fill = bar_colour, colour = "white") +
  labs(
    title = "Distribution of Gender",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal()
```

## Distribution of Smoking Status

```{r}
ggplot(df, aes(x = Smoking)) +
  geom_bar(fill = bar_colour, colour = "white") +
  labs(
    title = "Distribution of Smoking Status",
    x = "Smoking Status",
    y = "Count"
  ) +
  theme_minimal()
```

## Distribution of Age

```{r}
ggplot(df, aes(x = Age)) +
  geom_histogram(bins = 30, fill = histogram_colour, colour = "white") +
  labs(
    title = "Age Distribution",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Distribution of BMI

```{r}
ggplot(df, aes(x = BMI)) +
  geom_histogram(bins = 30, fill = histogram_colour, colour = "white") +
  labs(
    title = "BMI Distribution",
    x = "BMI",
    y = "Frequency"
  ) +
  theme_minimal()
```


## Age by Self-Reported Health

```{r}
ggplot(df, aes(x = Health, y = Age)) +
  geom_boxplot(fill = box_fill_colour, colour = box_border_colour) +
  labs(
    title = "Age by Self-Reported Health",
    x = "Health Category",
    y = "Age"
  ) +
  theme_minimal()
```

## BMI by Self-Reported Health

```{r}
ggplot(df, aes(x = Health, y = BMI)) +
  geom_boxplot(fill = box_fill_colour, colour = box_border_colour) +
  labs(
    title = "BMI by Self-Reported Health",
    x = "Health Category",
    y = "BMI"
  ) +
  theme_minimal()
```

# Parallel Cumulative Logit Model

```{r}
newdata_parallel <- data.frame(
  Age = seq(min(df$Age), max(df$Age), length.out = 100),
  Gender = factor(rep(levels(df$Gender)[1], 100), levels = levels(df$Gender)),
  BMI = rep(mean(df$BMI, na.rm = TRUE), 100),
  Smoking = factor(rep(levels(df$Smoking)[1], 100), levels = levels(df$Smoking))
)

head(newdata_parallel)
```

```{r}
parallel_model <- vglm(
  Health ~ Age + Gender + BMI + Smoking,
  family = cumulative(link = "logitlink", parallel = TRUE),
  data = df,
  model = TRUE
)

summary(parallel_model)
```

```{r}
pred_parallel <- predict(parallel_model, newdata = newdata_parallel, type = "response")

plot_parallel <- cbind(newdata_parallel, as.data.frame(pred_parallel)) %>%
  `colnames<-`(c("Age", "Gender", "BMI", "Smoking", "1", "2", "3", "4", "5")) %>%
  pivot_longer(
    cols = c("1", "2", "3", "4", "5"),
    names_to = "Health",
    values_to = "Probability"
  ) %>%
  mutate(
    Health = factor(
      Health,
      levels = c("1", "2", "3", "4", "5"),
      labels = c("Excellent", "Very good", "Good", "Fair", "Poor")
    ),
    Model = "Parallel"
  )

ggplot(plot_parallel, aes(x = Age, y = Probability, color = Health)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Parallel Cumulative Logit Model",
    x = "Age",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

# Non-Parallel Cumulative Logit Model

```{r}
nonparallel_model <- vglm(
  Health ~ Age + BMI + Gender + Smoking,
  family = cumulative(link = "logitlink", parallel = FALSE),
  data = df,
  model = TRUE
)

nonparallel_model
```

```{r}
newdata_nonparallel <- data.frame(
  Age = seq(min(df$Age), max(df$Age), length.out = 100),
  BMI = rep(mean(df$BMI, na.rm = TRUE), 100),
  Gender = factor(rep(levels(df$Gender)[1], 100), levels = levels(df$Gender)),
  Smoking = factor(rep(levels(df$Smoking)[1], 100), levels = levels(df$Smoking))
)

pred_nonparallel <- predictvglm(
  nonparallel_model,
  newdata = newdata_nonparallel,
  type = "response"
)

pred_nonparallel <- as.data.frame(pred_nonparallel)
colnames(pred_nonparallel) <- c("1", "2", "3", "4", "5")
```

```{r}
plot_nonparallel <- cbind(newdata_nonparallel, as.data.frame(pred_nonparallel)) %>%
  `colnames<-`(c("Age", "BMI", "Gender", "Smoking", "1", "2", "3", "4", "5")) %>%
  pivot_longer(
    cols = c("1", "2", "3", "4", "5"),
    names_to = "Health",
    values_to = "Probability"
  ) %>%
  mutate(
    Health = factor(
      Health,
      levels = c("1", "2", "3", "4", "5"),
      labels = c("Excellent", "Very good", "Good", "Fair", "Poor")
    ),
    Model = "Non-parallel"
  )

ggplot(plot_nonparallel, aes(x = Age, y = Probability, color = Health)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Non-Parallel Cumulative Logit Model",
    x = "Age",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

# Model Comparison

```{r}
AIC(parallel_model)
AIC(nonparallel_model)

BIC(parallel_model)
BIC(nonparallel_model)
```

```{r}
plot_both <- bind_rows(plot_parallel, plot_nonparallel)
```

# Interactive Model Comparison Graph

This section shows the previous model comparison idea as **two interactive graphs**: one for the non-parallel model and one for the parallel model. Hover over the lines to see the model, health category, age and predicted probability.

```{r}
plot_both <- bind_rows(plot_parallel, plot_nonparallel)

health_colours <- c(
  "Excellent" = "#F8766D",
  "Very good" = "#A3A500",
  "Good" = "#00BF7D",
  "Fair" = "#00B0F6",
  "Poor" = "#E76BF3"
)

make_model_comparison_plot <- function(model_name) {
  model_data <- plot_both %>% filter(Model == model_name)
  p <- plot_ly()

  for (h in levels(model_data$Health)) {
    temp <- model_data %>% filter(Health == h)

    p <- p %>%
      add_lines(
        data = temp,
        x = ~Age,
        y = ~Probability,
        name = h,
        legendgroup = h,
        showlegend = model_name == "Non-parallel",
        line = list(color = health_colours[[h]], width = 2),
        customdata = ~paste0(
          "Model: ", Model,
          "<br>Health Category: ", Health,
          "<br>Age: ", round(Age, 1),
          "<br>Predicted Probability: ", round(Probability, 4)
        ),
        hovertemplate = "%{customdata}<extra></extra>"
      )
  }

  p %>%
    layout(
      title = model_name,
      xaxis = list(title = "Age"),
      yaxis = list(title = "Predicted Probability")
    )
}

subplot(
  make_model_comparison_plot("Non-parallel"),
  make_model_comparison_plot("Parallel"),
  nrows = 1,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE
) %>%
  layout(
    title = "Interactive Model Comparison: Parallel vs Non-Parallel Cumulative Logit Models",
    legend = list(title = list(text = "Health Category"))
  )
```

# Interactive Health Category Prediction for a Particular Person

Enter a person's **Age**, **BMI**, **Gender** and **Smoking Status**.  
The graph shows which self-reported health category the person is most likely to fall into.

```{r}
health_labels <- c("Excellent", "Very good", "Good", "Fair", "Poor")

predict_person_parallel <- function(age, bmi, gender, smoking) {
  new_person <- data.frame(
    Age = age,
    Gender = factor(gender, levels = levels(df$Gender)),
    BMI = bmi,
    Smoking = factor(smoking, levels = levels(df$Smoking))
  )

  pred <- predict(parallel_model, newdata = new_person, type = "response")
  pred <- as.data.frame(pred)
  colnames(pred) <- health_labels

  pred %>%
    pivot_longer(
      cols = everything(),
      names_to = "Health",
      values_to = "Probability"
    ) %>%
  mutate(Health = factor(Health, levels = health_order, ordered = TRUE)) %>%
    mutate(
      Model = "Parallel",
      Probability = round(Probability, 4)
    )
}

predict_person_nonparallel <- function(age, bmi, gender, smoking) {
  new_person <- data.frame(
    Age = age,
    BMI = bmi,
    Gender = factor(gender, levels = levels(df$Gender)),
    Smoking = factor(smoking, levels = levels(df$Smoking))
  )

  pred <- predictvglm(nonparallel_model, newdata = new_person, type = "response")
  pred <- as.data.frame(pred)
  colnames(pred) <- health_labels

  pred %>%
    pivot_longer(
      cols = everything(),
      names_to = "Health",
      values_to = "Probability"
    ) %>%
  mutate(Health = factor(Health, levels = health_order, ordered = TRUE)) %>%
    mutate(
      Model = "Non-parallel",
      Probability = round(Probability, 4)
    )
}
```

```{r, echo=FALSE}
fluidRow(
  column(
    width = 3,
    numericInput(
      inputId = "input_age",
      label = "Age",
      value = 27,
      min = min(df$Age),
      max = max(df$Age),
      step = 1
    )
  ),
  column(
    width = 3,
    numericInput(
      inputId = "input_bmi",
      label = "BMI",
      value = 25,
      min = floor(min(df$BMI, na.rm = TRUE)),
      max = ceiling(max(df$BMI, na.rm = TRUE)),
      step = 0.1
    )
  ),
  column(
    width = 3,
    selectInput(
      inputId = "input_gender",
      label = "Gender",
      choices = levels(df$Gender),
      selected = levels(df$Gender)[1]
    )
  ),
  column(
    width = 3,
    selectInput(
      inputId = "input_smoking",
      label = "Smoking Status",
      choices = levels(df$Smoking),
      selected = levels(df$Smoking)[1]
    )
  )
)
```

```{r, echo=FALSE}
person_predictions <- reactive({
  parallel_result <- predict_person_parallel(
    age = input$input_age,
    bmi = input$input_bmi,
    gender = input$input_gender,
    smoking = input$input_smoking
  )

  nonparallel_result <- predict_person_nonparallel(
    age = input$input_age,
    bmi = input$input_bmi,
    gender = input$input_gender,
    smoking = input$input_smoking
  )

  bind_rows(parallel_result, nonparallel_result) %>%
    group_by(Model) %>%
    mutate(
      Highest_Probability = Probability == max(Probability),
      Hover_Text = paste0(
        "Model: ", Model,
        "<br>Health Category: ", Health,
        "<br>Predicted Probability: ", Probability,
        ifelse(Highest_Probability, "<br><b>Predicted Category</b>", "")
      )
    ) %>%
    ungroup()
})
```

## Graph Showing the Predicted Health Category

This graph shows the predicted probabilities for each health category. Nothing is written inside or on top of the bars. Move the mouse over a bar to see the model, category, probability and predicted category.

```{r, echo=FALSE}
renderPlotly({
  pink_data <- person_predictions() %>% filter(Model == "Parallel")
  grey_data <- person_predictions() %>% filter(Model == "Non-parallel")

  plot_ly() %>%
    add_bars(
      data = pink_data,
      x = ~Health,
      y = ~Probability,
      name = "Parallel",
      marker = list(color = "pink"),
      customdata = ~Hover_Text,
      hovertemplate = "%{customdata}<extra></extra>"
    ) %>%
    add_bars(
      data = grey_data,
      x = ~Health,
      y = ~Probability,
      name = "Non-parallel",
      marker = list(color = "grey"),
      customdata = ~Hover_Text,
      hovertemplate = "%{customdata}<extra></extra>"
    ) %>%
    layout(
      title = "Which Health Category Does This Person Fall Into?",
      barmode = "group",
      xaxis = list(title = "Health Category", categoryorder = "array", categoryarray = health_order),
      yaxis = list(title = "Predicted Probability", range = c(0, 1)),
      legend = list(title = list(text = "Model"))
    )
})
```

## Final Predicted Category

```{r, echo=FALSE}
renderTable({
  person_predictions() %>%
    filter(Highest_Probability == TRUE) %>%
    dplyr::select(Model, Predicted_Category = Health, Probability)
})
```

## Full Probability Table

```{r, echo=FALSE}
renderTable({
  person_predictions() %>%
    dplyr::select(Model, Health, Probability)
})
```
