---
title: "MSc Health Data Science Project Title"
author:
  - name: "Manasvi Saite"
    affil: 1
affiliation:
  - num: 1
    address: "MSc Health Data Science, University of Galway"
logoleft_name: "/Users/manasvi15/Downloads/Project/UOG_logo.png"

# A1 Landscape
poster_height: "594mm"
poster_width: "841mm"

# Layout
column_numbers: 4
column_margins: "0.5in"

# Colours
primary_colour: "#a80050"
secondary_colour: "#84003d"
accent_colour: "#e6007e"

# Text sizes
author_textcol: "white"
title_textsize: "90pt"
author_textsize: "60pt"
body_textsize: "25pt"

output:
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false

# Optional: creates PDF via Chrome when you Knit (comment out if you only want HTML)
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{ width: 20%; }
div.poster_title{ width: 80%; }
.section h4 { break-after: column; }
div.footnotes { font-size: 18pt; }
```

<!--
This is the GENERAL POSTER TEMPLATE.
Step 1: Knit to HTML to see the poster layout.
Step 2: Replace the placeholder text with your project content.
Step 3: Add your tables/figures in the chunks below.
-->

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)
options(knitr.table.format = "html")

library(tidyverse)
library(kableExtra)
library(plotly)   # optional (for interactive plots in HTML)
# library(readxl) # if your dataset is Excel
```

# Background and Problem

Write 3–5 lines:
- What is the problem?
- Why is it important in health/clinical/public health context?
- What is the gap you are addressing?

[^1]: Example reference style (footnote). Replace with your own sources.

# Objectives of Project

1. Objective 1 (one short sentence)
2. Objective 2
3. Objective 3

# Data Sources and Datasets

**Dataset:** (name)  
**Source:** (where it came from)  
**Sample size:** (n = )  
**Key variables:** (list 4–8 variables)

```{r data_load, include=FALSE}
# ===== LOAD YOUR DATA HERE =====
# Put your file in a folder called "data" in the same directory as this Rmd.

# CSV:
# df <- readr::read_csv("data/your_dataset.csv")

# Excel:
# df <- readxl::read_excel("data/your_dataset.xlsx")

# Quick check:
# glimpse(df)
```

####

# Early Results / Descriptive Statistics of Datasets

## Table 1: Descriptive summary (replace with your own)

Tip: Keep tables short and readable (A1 poster = big font, less rows).

```{r table1, out.width='95%'}
# Example (REPLACE with your columns):
# table1_df <- df %>%
#   summarise(
#     n = n(),
#     age_mean = mean(age, na.rm = TRUE),
#     age_sd   = sd(age, na.rm = TRUE),
#     male_pct = mean(sex == "Male", na.rm = TRUE) * 100
#   )
#
# knitr::kable(table1_df, caption = "Table 1: Summary statistics", align = "c") %>%
#   kable_styling(full_width = FALSE, font_size = 20)
```

## Figure 1: (replace caption + code)

```{r fig1, out.width='90%', fig.cap='Figure 1: Your caption here', fig.height=4}
# Example histogram (REPLACE):
# ggplot(df, aes(x = age)) +
#   geom_histogram(bins = 30) +
#   theme_minimal(base_size = 22) +
#   labs(x = "Age", y = "Count")
```

## Figure 2: (replace caption + code)

```{r fig2, out.width='90%', fig.cap='Figure 2: Your caption here', fig.height=4}
# Example relationship plot (REPLACE):
# ggplot(df, aes(x = bmi, y = outcome)) +
#   geom_point(alpha = 0.5) +
#   geom_smooth(method = "lm", se = TRUE) +
#   theme_minimal(base_size = 22) +
#   labs(x = "BMI", y = "Outcome")
```

## Optional: Interactive plot (HTML only)

```{r interactive, out.width='90%', fig.cap='Figure 3: Interactive plot (optional)', fig.height=4}
# Example (REPLACE):
# p <- ggplot(df, aes(x = age, y = bmi)) +
#   geom_point(alpha = 0.5) +
#   theme_minimal(base_size = 22)
# plotly::ggplotly(p)
```

####

# Next Project Steps

- Next step 1 (e.g., data cleaning / feature engineering)
- Next step 2 (e.g., modelling: logistic regression / survival / ML)
- Next step 3 (e.g., validation / sensitivity analysis)

# GitHub

(Optional) Add your repository link:  
<https://github.com/yourusername/your-repo>

# References

Add references as footnotes like this:  
Example: This method is widely used in healthcare analytics.[^2]

[^2]: Author, Title, Journal, Year.
