---
title: "Count Data On Health Industry"
author:
  - name: "Manasvi Saite & Kaushal Mayur"
affiliation:
    address: "MSc Health Data Science, University of Galway"
logoleft_name: "/Users/manasvi15/Downloads/Project/UOG_logo.png"

# A1 Landscape
poster_height: "594mm"
poster_width: "841mm"

# Layout
column_numbers: 4
column_margins: "0.5in"

# Colours
primary_colour: "#a80050"
secondary_colour: "#84003d"
accent_colour: "#e6007e"

# Text sizes
author_textcol: "white"
title_textsize: "90pt"
author_textsize: "60pt"
body_textsize: "25pt"

output:
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false

# Optional: creates PDF via Chrome when you Knit (comment out if you only want HTML)
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{ width: 20%; }
div.poster_title{ width: 80%; }
.section h4 { break-after: column; }
div.footnotes { font-size: 18pt; }
```

<!--
This is the GENERAL POSTER TEMPLATE.
Step 1: Knit to HTML to see the poster layout.
Step 2: Replace the placeholder text with your project content.
Step 3: Add your tables/figures in the chunks below.
-->

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)
options(knitr.table.format = "html")

library(tidyverse)
library(kableExtra)
library(plotly)   # optional (for interactive plots in HTML)
# library(readxl) # if your dataset is Excel
```

# Background and Objective

Healthcare systems routinely generate count data, such as the number of hospital admissions.
This project focuses on analysing hospitalisation counts based on disease type, sex, and other patient-related factors.

Hospitalisation data consist of non-negative integers and frequently display overdispersion and a high number of zero admissions, where many individuals experience no hospitalisations during the study period. These characteristics make standard regression methods unsuitable.

To address this, count data models provide an appropriate statistical framework for modelling hospital admissions and understanding patterns of healthcare utilisation.

Objectives of Project

To analyse hospitalisation counts across different disease groups.

To examine the effect of sex and other patient characteristics on hospitalisation rates.

To apply suitable count data models for analysing hospital admission data.

# Data Sources and Datasets

**Dataset:** (name)  
**Source:** (where it came from)  
**Sample size:** (n = )  
**Key variables:** (list 4–8 variables)

```{r data_load, include=FALSE}
# ===== LOAD YOUR DATA HERE =====
# Put your file in a folder called "data" in the same directory as this Rmd.

# CSV:
# df <- readr::read_csv("data/your_dataset.csv")

# Excel:
# df <- readxl::read_excel("data/your_dataset.xlsx")

# Quick check:
# glimpse(df)
```

####

# Early Results / Descriptive Statistics of Datasets

## Table 1: Descriptive summary (replace with your own)

Tip: Keep tables short and readable (A1 poster = big font, less rows).

```{r table1, out.width='95%'}
# Example (REPLACE with your columns):
# table1_df <- df %>%
#   summarise(
#     n = n(),
#     age_mean = mean(age, na.rm = TRUE),
#     age_sd   = sd(age, na.rm = TRUE),
#     male_pct = mean(sex == "Male", na.rm = TRUE) * 100
#   )
#
# knitr::kable(table1_df, caption = "Table 1: Summary statistics", align = "c") %>%
#   kable_styling(full_width = FALSE, font_size = 20)
```

## Figure 1: (replace caption + code)

```{r fig1, out.width='90%', fig.cap='Figure 1: Your caption here', fig.height=4}
# Example histogram (REPLACE):
# ggplot(df, aes(x = age)) +
#   geom_histogram(bins = 30) +
#   theme_minimal(base_size = 22) +
#   labs(x = "Age", y = "Count")
```

## Figure 2: (replace caption + code)

```{r fig2, out.width='90%', fig.cap='Figure 2: Your caption here', fig.height=4}
# Example relationship plot (REPLACE):
# ggplot(df, aes(x = bmi, y = outcome)) +
#   geom_point(alpha = 0.5) +
#   geom_smooth(method = "lm", se = TRUE) +
#   theme_minimal(base_size = 22) +
#   labs(x = "BMI", y = "Outcome")
```

## Optional: Interactive plot (HTML only)

```{r interactive, out.width='90%', fig.cap='Figure 3: Interactive plot (optional)', fig.height=4}
# Example (REPLACE):
# p <- ggplot(df, aes(x = age, y = bmi)) +
#   geom_point(alpha = 0.5) +
#   theme_minimal(base_size = 22)
# plotly::ggplotly(p)
```

####

# Next Project Steps

- Next step 1 (e.g., data cleaning / feature engineering)
- Next step 2 (e.g., modelling: logistic regression / survival / ML)
- Next step 3 (e.g., validation / sensitivity analysis)

# GitHub

(Optional) Add your repository link:  
<https://https://github.com/manasvisaite1504/Count_data_health_Industry-project>

# References

Add references as footnotes like this:  
Example: This method is widely used in healthcare analytics.[^2]

[^2]: Author, Title, Journal, Year.
